---
title: "Practical Machine Learning 2014"
author: "Costas Voglis"
date: "Friday, December 05, 2014"
output: html_document
---
#Programming Assignement - Weight Lifting Exercise Dataset

In this assignement the task was to predict a weight lifting exercise class using
data from accelerometers, attached on the body of 6 individuals. Each class
represents a different type of exercise. We will divide the process in two phases

1. Data preprocessing: Perform analysis on the columns of the original data and keep only the meaningful ones
2. Model selection: Perform parameter analysis on various well known classifiers

## 1.Data preprocessing
Initialy we load the data using: 
```{r, echo=TRUE}
library(caret)
dataOriginal<-read.csv("pml-training.csv")
````

A quick look in the dimensionality of the data frame reveals that we have 19622 observations
with 160 columns.
```{r, echo=TRUE}
dim(dataOriginal)
```

We can discard the first 5 columns since they represent user data and timestamps
```{r, echo=TRUE}
head(dataOriginal[, (1:5)])
dataOriginal<-dataOriginal[, -(1:5)]
```

Find near zero variance indices using caret's  *nearZeroVar* with the default values.
```{r, echo=TRUE}
indexNearZero <- nearZeroVar(dataOriginal)
str(indexNearZero)
data<-dataOriginal[, -indexNearZero]
dim(data)
```

We have reduced the number of columns to `r dim(data)[2]`.

Next we notice that many remaining predictors have large number of NA values
```{r, echo=TRUE}
NAdata<-is.na(data)
NAdataCount<-colSums(NAdata)
```

First We find these predictrors:
```{r, echo=TRUE}
LargeNAdataCount<-NAdataCount[NAdataCount>19000]
LargeNAdataCountNames<-names(LargeNAdataCount)
LargeNAdataCountNames
```

And we exclude these predictors them: 

```{r, echo=TRUE}
data<-data[, -which(names(data) %in% LargeNAdataCountNames)]
```

At this point we have `r dim(data)[2]` predictors left. Finally we perform
a correlation analysis among the remaining predictors using caret's findCorrelation.
In this setting a cut off-threshold of 80% was used.
We exclude column 54 (classe).

```{r, echo=TRUE}
correlated <- findCorrelation(cor(data[, -54]), cutoff = 0.80, verbose=FALSE)
data<-data[, -correlated]
dim(data)
```

We are now ready to perfom model training and selection. The final dataset consists of `r dim(data)[1]` samples
having `r dim(data)[2]-1` numeric predictors. 

```{r}
str(data)
````

## 2.Model selection

In this study we use SVM and Random Forest classifiers since it has been shown
that they produce accurate models with good generalization properties.

1. SVM classifier
```{r}
#classModel_RF <-train(x=data[, -42], y=data[, 42], method="svmRadial", trControl = trainControl(method = "cv", number=4, allowParallel = TRUE))
```

2. Random forests
```{r}
#classModel_RF <-train(x=data[, -42], y=data[, 42], method="rf", trControl = trainControl(method = "cv", number=4, allowParallel = TRUE))



