---
title: "Practical Machine Learning 2014"
author: "Costas Voglis"
date: "Friday, December 05, 2014"
output: html_document
---
#Programming Assignement - Weight Lifting Exercise Dataset

In this assignement the task was to predict a weight lifting exercise class using
data from accelerometers, attached on the body of 6 individuals. Each class
represents a different type of exercise. We will divide the process in two phases

1. Data preprocessing: Perform analysis on the columns of the original data and keep only the meaningful ones
2. Model selection: Perform parameter analysis on various well known classifiers

## 1.Data preprocessing
Initialy we load the data using: 
```{r, echo=TRUE}
library(caret)
dataOriginal<-read.csv("pml-training.csv")
````

A quick look in the dimensionality of the data frame reveals that we have 19622 observations
with 160 columns.
```{r, echo=TRUE}
dim(dataOriginal)
```

We can discard the first 5 columns since they represent user data and timestamps
```{r, echo=TRUE}
head(dataOriginal[, (1:5)])
dataOriginal<-dataOriginal[, -(1:5)]
```

Find near zero variance indices using caret's  *nearZeroVar* with the default values.
```{r, echo=TRUE}
indexNearZero <- nearZeroVar(dataOriginal)
str(indexNearZero)
data<-dataOriginal[, -indexNearZero]
dim(data)
```

We have reduced the number of columns to `r dim(data)[2]`.


## 2.Model selection



